Inverse energy transfer in three-dimensional quantum vortex flows

### 2025-06-26 main_REV.tex
FInalised figures
Removed coloured text

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{blindtext}
\usepackage{float}
\usepackage{caption,cancel}
\usepackage{cleveref}
\usepackage{subcaption}
\usepackage{xcolor}
%\usepackage{hyperref}% add hypertext capabilities
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

	% \end{subfigure} \hfill
  %   %
	% \begin{subfigure}[b]{0.24\textwidth}
	% 	\centering
	% 	\includegraphics*[width=\textwidth]{snaps-hel-1.pdf}
	% \end{subfigure}
	% \begin{subfigure}[b]{0.24\textwidth}
	% 	\centering
	% 	\includegraphics*[width=\textwidth]{snaps-hel-2.pdf}
	% \end{subfigure}
  %   \begin{subfigure}[b]{0.24\textwidth}
	% 	\centering
	% 	\includegraphics*[width=\textwidth]{snaps-hel-3.pdf}
	% \end{subfigure}
  %   \begin{subfigure}[b]{0.24\textwidth}
	% 	\centering
	% 	\includegraphics*[width=\textwidth]{snaps-hel-4.pdf}
	% \end{subfigure} \hfill
	\caption{
Three-dimensional rendering of the time evolution of
an initially orthogonal vortex configuration 
undergoing a vortex reconnection at $T = 1.9$K at dimensionless
times (from left to right)
$(t-t_R)/\tau_R=-7.47\times10^{-2}$, $-3.09\times10^{-3}$, $8.79\times10^{-4}$ and $5.39\times10^{-3}$, where $\tau_R = L_R/E^{1/2}_R$. The quantities $E_R$ and $L_R$ are the energy $E$ and integral length scales \red{$L$} of the normal fluid at reconnection time $t_R$. The green
tubes represent the superfluid vortex lines (the tubes’ radii have been 
greatly exaggerated for visual purpose). In the top sequence,
the blue volume rendering represents the scaled normal fluid enstrophy 
$\bom^2/\bom^2_{max}$. Note the Kelvin wave on the superfluid
vortex at $(t-t_R)/\tau_R=8.79\times10^{-4}$ . In the bottom sequence, the red/blue volume
rendering at the same times represent scaled positive/negative normal fluid
helicity.	
%Three-dimensional rendering of an orthogonal vortex configuration 
%undergoing a vortex reconnection at $T=1.9K$. The red tubes represent
%the superfluid vortex
%lines (the tubes' radii have been greatly exaggerated for visual purposes), 
%and the blue volume rendering represents the scaled normal fluid enstrophy 
%$\bom^2/\bom^2_{max}$. In the third panel, note the Kelvin waves on the superfluid vortices.
}
    \label{fig:visualisation}
\end{figure*}

@@

Remarkably, the same cascade phenomenology is observed in turbulent flows 
of quantum fluids, {\textit{i.e.} fluids at very low temperatures whose physics is
dominated by quantum effects.
Examples of such fluids are superfluid helium 
and atomic Bose-Einstein Condensates (BECs). 
The dynamics of these systems can be successfully depicted in terms of 
a two-fluid model \cite{tisza-1938,landau-1949,skrbek-sreenivasan-2012} 
describing the quantum fluid as the mixture of two components, 
the superfluid component and the thermal (or normal) component, which 
interact by means of a mutual friction force 
\cite{jackson-etal-2009,hall-vinen-1956a,hall-vinen-1956b}. 
The superfluid component flows without viscosity and
vanishing entropy; its vorticity is
confined to effectively one-dimensional vortex filaments
of atomic core thickness (called quantum vortices or vortex lines), 
around which the circulation of the velocity is quantised.
In BECs the thermal component forms a ballistic gas,
whereas in superfluid $^4$He it can be described as
a classical viscous fluid.
Despite these significant differences with respect to ordinary fluids, 
the direct
%forward 
kinetic energy cascade has
indeed been observed in three-dimensional superfluid
turbulence 
\cite{maurer1998,salort2010turbulent,baggaley2012,sherwin-robson2015,Muller_KolmogorovKelvinWave_2020,Muller_IntermittencyVelocityCirculation_2021}.
Evidence of this 
%forward 
direct
cascade has been found also in
three-dimensional turbulent BECs \cite{middleton-spencer2022}. 
\red{
In this forward energy cascade a fundamental role is played by the dynamics and interactions of quantum vortices,
in particular by their reconnections, phenomena where  
two vortex lines collide and recombine, exchanging heads and tails, 
altering the overall topology of the flow
\cite{koplik-levine-1993,bewley-etal-2008,rorai-etal-2016,serafini-etal-2017,galantucci-baggaley-parker-barenghi-2019,villoisUniversalNonuniversalAspects2017,villois2020irreversible}.
Reconnections occur frequently in 3D quantum turbulence and are essential to
the energy transfer towards small scales by engendering the the breakdown of coherent vortex structures
and triggering the Kelvin-wave cascade \cite{vinen-2001}.
}
In 2D, similarly to classical turbulence, 
an inverse energy cascade characterises turbulence in two-dimensional BECs, 
as shown in theoretical \cite{bradley2012energy,reeves2013,simula2014emergence,Muller_ExploringEquivalenceTwoDimensional_2024} and experimental \cite{johnstone2019evolution,gauthier2019giant} studies.

@@

To study the reconnection dynamics, we consider two pairs of initially orthogonal vortices (where the corresponding vortices of each pair have opposite circulation in order to preserve periodicity along the boundaries)
at two distinct temperatures, $T=1.9K$ and $T=2.1K$. The vortex pairs are separated by the distance $D_{\ell}$; each vortex within each pair is initially at distance $d_{\ell}$ to the other vortex, such that $d_{\ell}\ll D_{\ell}$ to ensures that the dynamics in the vicinity of the reconnection is dominated by local interactions, and that the far-field contribution from the other vortex pair is negligible. 
\red{All results reported hereafter refer only to the reconnection of one pair of vortices.} 
%\red{In our simulations, the reconnections of the two pairs of vortices occur at different times; here we consider only the vortex dynamics up to and shortly after the first reconnection (see Fig.~\ref{fig:total-helicity}). Hence, all results reported and discussed herafter refer to the first reconnection.}.

@@

A chiral imbalance occurs if the mutual friction force is helical, 
\textit{i.e.} if the ratio $|f^+|^2/|f^-|^2 \neq 1$, with $|f^\pm|^2$ the total squared norm of the mutual friction components. 
In Fig.~\ref{fig:mutual-friction-decomp}, we show the temporal evolution of $|f^+|^2/|f^-|^2$, for both temperatures. 
%
\begin{figure}[h!]
    \centering
    \includegraphics*[width=0.48\textwidth]{fmfDecompFig.pdf}
    \caption{Temporal evolution of projected mutual friction force components $f^\pm$/ Inset: temporal evolution of total helical components.}.
    \label{fig:mutual-friction-decomp}
\end{figure}
%
It is apparent
that during and after the reconnection, the mutual friction force is strongly chiral,
injecting more negative helicity than positive helicity. 
As a result, the ratio $\mathcal{H}^+/\mathcal{H}^-$ 
(reported in the inset of Fig.~\ref{fig:mutual-friction-decomp}) 
decreases significantly at reconnection and remains smaller than unity even at later times, 
indicating that the flow is persistently chiral. We conclude that the reconnection triggers indeed a chiral imbalance.
\red{
For the sake of completeness, it is worth noting that the total helicity of the flow (obtained by adding the superfluid helicity to that of the normal fluid $\mathcal{H}$) is not conserved.
}
From Fig.~\ref{fig:mutual-friction-decomp}, \red{ we consider a time $t^*$ after reconnection where the ratio $|f^+|^2/|f^-|^2$ has sufficiently decayed to a roughly 5\% difference and} we determine the non-dimensional timescale \red{$\tau = (t^*-t_R)/\tau_R$} during which the mutual friction force is chiral as a result of reconnections: $\tau \approx 0.01$ and $\tau \approx 0.005$
for $T=1.9$K and $T=2.1$K, respectively, corresponding dimensionally to $\tau \approx 0.1$s for both temperatures. 
In superfluid turbulence, the timescale between two consecutive reconnections can be smaller than $\tau$ provided that the
vortex line density $\mathcal{L}$ (length of vortices per unit volume) is larger than $10^8 \text{m}^{-2}$ \cite{stasiak2025experimental,barenghi2004},
a condition which is easily met in superfluid helium experiments \cite{roche2007,Babuin2014}.

@@

%\paragraph*{Closing remarks.---}
In conclusion, the reconnection of quantum vortices in the two-fluid regime
($T\gtrsim 1.5$K) not only injects punctuated
energy in the normal fluid \cite{stasiak2025experimental}, but also triggers in the normal fluid
a transfer of kinetic energy towards the large scales. 
%This inverse 
%energy transfer arises from the fact that the mutual friction force (injecting energy and 
%helicity in the normal fluid) is helical, as Kelvin waves develop on the vortices as a result of the reconnection itself.
%The helical character of the mutual friction produces a chiral imbalance in the normal fluid, driving this inverse transfer, as previously observed in turbulent Navier-Stokes flows
This inverse energy transfer \red{is triggered by} the helical character of the
mutual friction, \red{directly arising from the} Kelvin waves released by the reconnecting
cusp, which produces a chiral imbalance in the normal fluid, as 
previously observed in turbulent Navier-Stokes flows
\cite{biferaleInverseEnergyCascade2012a,plunianInverseCascadeEnergy2020a}.

Our findings have profound implications for the nature of turbulence in 
finite temperature superfluids, \red{where vortex reconnections play a key role}. 
In circumstances where the vortex density 
$\mathcal{L}$ is large and where the isotropy of the vortex tangle 
is broken by external forcing (for instance thermal counterflows~\cite{vinen-1957a}), 
the chirality of the flow generated by the 
frequent \red{non-symmetrical} reconnections may be strong enough to induce an inverse energy cascade~\cite{notes_recon}.
%The asymmetry of the reconnecting vortex configurations is a key element; 
%when perfectly symmetrical vortex configurations (with corresponding 
%opposite vortex line orientations) reconnect, the injected helicity 
%globally cancels out, as verified in our study if we take also into 
%account the reconnection.
The microscopic mechanism that we have described \red{may be relevant in triggering}
the inverse energy cascade which is observed numerically in  
large-scale simulations of counterflow turbulence at large heat fluxes \cite{polanco2020}, which is indeed inherently not isotropic. \magenta{However, formalising the connection between high counterflow, unbalanced helicity injection and inverse energy cascade is  challenging, as the model used in \cite{polanco2020} and the one used in our work describe superfluid helium at very different scales.}
%, hence triggering non-fully reconnections. 
Our work hence motivates further detailed studies of
the role played by helicity in superfluid dynamics 
\cite{dileoni2016,galantucci2021},
moving the emphasis from few vortex systems \cite{scheeler2014} 
to fully coupled superfluid turbulence.

>>>>

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal poi<?xml version="1.0" encoding="UTF-8" standalone="no"?>
\usepackage{bm}% bold math
\usepackage{blindtext}
\usepackage{float}
\usepackage{caption,cancel}
\usepackage{cleveref}
\usepackage{subcaption}
\usepackage{xcolor}
%\usepackage{hyperref}% add hypertext capabilities
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

	% \end{subfigure} \hfill
  %   %
	% \begin{subfigure}[b]{0.24\textwidth}
	% 	\centering
	% 	\includegraphics*[width=\textwidth]{snaps-hel-1.pdf}
	% \end{subfigure}
	% \begin{subfigure}[b]{0.24\textwidth}
	% 	\centering
	% 	\includegraphics*[width=\textwidth]{snaps-hel-2.pdf}
	% \end{subfigure}
  %   \begin{subfigure}[b]{0.24\textwidth}
	% 	\centering
	% 	\includegraphics*[width=\textwidth]{snaps-hel-3.pdf}
	% \end{subfigure}
  %   \begin{subfigure}[b]{0.24\textwidth}
	% 	\centering
	% 	\includegraphics*[width=\textwidth]{snaps-hel-4.pdf}
	% \end{subfigure} \hfill
	\caption{
Three-dimensional rendering of the time evolution of
an initially orthogonal vortex configuration 
undergoing a vortex reconnection at $T = 1.9$K at dimensionless
times (from left to right)
$(t-t_R)/\tau_R=-7.47\times10^{-2}$, $-3.09\times10^{-3}$, $8.79\times10^{-4}$ and $5.39\times10^{-3}$, where $\tau_R = L_R/E^{1/2}_R$. The quantities $E_R$ and $L_R$ are the energy $E$ and integral length scales $L$ of the normal fluid at reconnection time $t_R$. The green
tubes represent the superfluid vortex lines (the tubes’ radii have been 
greatly exaggerated for visual purpose). In the top sequence,
the blue volume rendering represents the scaled normal fluid enstrophy 
$\bom^2/\bom^2_{max}$. Note the Kelvin wave on the superfluid
vortex at $(t-t_R)/\tau_R=8.79\times10^{-4}$ . In the bottom sequence, the red/blue volume
rendering at the same times represent scaled positive/negative normal fluid
helicity.	
%Three-dimensional rendering of an orthogonal vortex configuration 
%undergoing a vortex reconnection at $T=1.9K$. The red tubes represent
%the superfluid vortex
%lines (the tubes' radii have been greatly exaggerated for visual purposes), 
%and the blue volume rendering represents the scaled normal fluid enstrophy 
%$\bom^2/\bom^2_{max}$. In the third panel, note the Kelvin waves on the superfluid vortices.
}
    \label{fig:visualisation}
\end{figure*}

@@

Remarkably, the same cascade phenomenology is observed in turbulent flows 
of quantum fluids, {\textit{i.e.} fluids at very low temperatures whose physics is
dominated by quantum effects.
Examples of such fluids are superfluid helium 
and atomic Bose-Einstein Condensates (BECs). 
The dynamics of these systems can be successfully depicted in terms of 
a two-fluid model \cite{tisza-1938,landau-1949,skrbek-sreenivasan-2012} 
describing the quantum fluid as the mixture of two components, 
the superfluid component and the thermal (or normal) component, which 
interact by means of a mutual friction force 
\cite{jackson-etal-2009,hall-vinen-1956a,hall-vinen-1956b}. 
The superfluid component flows without viscosity and
vanishing entropy; its vorticity is
confined to effectively one-dimensional vortex filaments
of atomic core thickness (called quantum vortices or vortex lines), 
around which the circulation of the velocity is quantised.
In BECs the thermal component forms a ballistic gas,
whereas in superfluid $^4$He it can be described as
a classical viscous fluid.
Despite these significant differences with respect to ordinary fluids, 
the direct
%forward 
kinetic energy cascade has
indeed been observed in three-dimensional superfluid
turbulence 
\cite{maurer1998,salort2010turbulent,baggaley2012,sherwin-robson2015,Muller_KolmogorovKelvinWave_2020,Muller_IntermittencyVelocityCirculation_2021}.
Evidence of this 
%forward 
direct
cascade has been found also in
three-dimensional turbulent BECs \cite{middleton-spencer2022}. 
In this forward energy cascade a fundamental role is played by the dynamics and interactions of quantum vortices,
in particular by their reconnections, phenomena where  
two vortex lines collide and recombine, exchanging heads and tails, 
altering the overall topology of the flow
\cite{koplik-levine-1993,bewley-etal-2008,rorai-etal-2016,serafini-etal-2017,galantucci-baggaley-parker-barenghi-2019,villoisUniversalNonuniversalAspects2017,villois2020irreversible}.
Reconnections occur frequently in 3D quantum turbulence and are essential to
the energy transfer towards small scales by engendering the the breakdown of coherent vortex structures
and triggering the Kelvin-wave cascade \cite{vinen-2001}.
In 2D, similarly to classical turbulence, 
an inverse energy cascade characterises turbulence in two-dimensional BECs, 
as shown in theoretical \cite{bradley2012energy,reeves2013,simula2014emergence,Muller_ExploringEquivalenceTwoDimensional_2024} and experimental \cite{johnstone2019evolution,gauthier2019giant} studies.

@@

To study the reconnection dynamics, we consider two pairs of initially orthogonal vortices (where the corresponding vortices of each pair have opposite circulation in order to preserve periodicity along the boundaries)
at two distinct temperatures, $T=1.9K$ and $T=2.1K$. The vortex pairs are separated by the distance $D_{\ell}$; each vortex within each pair is initially at distance $d_{\ell}$ to the other vortex, such that $d_{\ell}\ll D_{\ell}$ to ensures that the dynamics in the vicinity of the reconnection is dominated by local interactions, and that the far-field contribution from the other vortex pair is negligible. 
All results reported hereafter refer only to the reconnection of one pair of vortices. 
%\red{In our simulations, the reconnections of the two pairs of vortices occur at different times; here we consider only the vortex dynamics up to and shortly after the first reconnection (see Fig.~\ref{fig:total-helicity}). Hence, all results reported and discussed herafter refer to the first reconnection.}.

@@

A chiral imbalance occurs if the mutual friction force is helical, 
\textit{i.e.} if the ratio $|f^+|^2/|f^-|^2 \neq 1$, with $|f^\pm|^2$ the total squared norm of the mutual friction components. 
In Fig.~\ref{fig:mutual-friction-decomp}, we show the temporal evolution of $|f^+|^2/|f^-|^2$, for both temperatures. 
%
\begin{figure}[h!]
    \centering
    \includegraphics*[width=0.48\textwidth]{fmfDecompFig.pdf}
    \caption{Temporal evolution of projected mutual friction force components $f^\pm$/ Inset: temporal evolution of total helical components.}.
    \label{fig:mutual-friction-decomp}
\end{figure}
%
It is apparent
that during and after the reconnection, the mutual friction force is strongly chiral,
injecting more negative helicity than positive helicity. 
As a result, the ratio $\mathcal{H}^+/\mathcal{H}^-$ 
(reported in the inset of Fig.~\ref{fig:mutual-friction-decomp}) 
decreases significantly at reconnection and remains smaller than unity even at later times, 
indicating that the flow is persistently chiral. We conclude that the reconnection triggers indeed a chiral imbalance.
For the sake of completeness, it is worth noting that the total helicity of the flow (obtained by adding the superfluid helicity to that of the normal fluid $\mathcal{H}$) is not conserved.
From Fig.~\ref{fig:mutual-friction-decomp}, we consider a time $t^*$ after reconnection where the ratio $|f^+|^2/|f^-|^2$ has sufficiently decayed to a roughly 5\% difference and we determine the non-dimensional timescale $\tau = (t^*-t_R)/\tau_R$ during which the mutual friction force is chiral as a result of reconnections: $\tau \approx 0.01$ and $\tau \approx 0.005$
for $T=1.9$K and $T=2.1$K, respectively, corresponding dimensionally to $\tau \approx 0.1$s for both temperatures. 
In superfluid turbulence, the timescale between two consecutive reconnections can be smaller than $\tau$ provided that the
vortex line density $\mathcal{L}$ (length of vortices per unit volume) is larger than $10^8 \text{m}^{-2}$ \cite{stasiak2025experimental,barenghi2004},
a condition which is easily met in superfluid helium experiments \cite{roche2007,Babuin2014}.

@@

%\paragraph*{Closing remarks.---}
In conclusion, the reconnection of quantum vortices in the two-fluid regime
($T\gtrsim 1.5$K) not only injects punctuated
energy in the normal fluid \cite{stasiak2025experimental}, but also triggers in the normal fluid
a transfer of kinetic energy towards the large scales. 
%This inverse 
%energy transfer arises from the fact that the mutual friction force (injecting energy and 
%helicity in the normal fluid) is helical, as Kelvin waves develop on the vortices as a result of the reconnection itself.
%The helical character of the mutual friction produces a chiral imbalance in the normal fluid, driving this inverse transfer, as previously observed in turbulent Navier-Stokes flows
This inverse energy transfer is triggered by the helical character of the
mutual friction, directly arising from the Kelvin waves released by the reconnecting
cusp, which produces a chiral imbalance in the normal fluid, as 
previously observed in turbulent Navier-Stokes flows
\cite{biferaleInverseEnergyCascade2012a,plunianInverseCascadeEnergy2020a}.

Our findings have profound implications for the nature of turbulence in 
finite temperature superfluids, where vortex reconnections play a key role. 
In circumstances where the vortex density 
$\mathcal{L}$ is large and where the isotropy of the vortex tangle 
is broken by external forcing (for instance thermal counterflows~\cite{vinen-1957a}), 
the chirality of the flow generated by the 
frequent non-symmetrical reconnections may be strong enough to induce an inverse energy cascade~\cite{notes_recon}.
%The asymmetry of the reconnecting vortex configurations is a key element; 
%when perfectly symmetrical vortex configurations (with corresponding 
%opposite vortex line orientations) reconnect, the injected helicity 
%globally cancels out, as verified in our study if we take also into 
%account the reconnection.
The microscopic mechanism that we have described may be relevant in triggering
the inverse energy cascade which is observed numerically in  
large-scale simulations of counterflow turbulence at large heat fluxes \cite{polanco2020}, which is indeed inherently not isotropic. However, formalising the connection between high counterflow, unbalanced helicity injection and inverse energy cascade is  challenging, as the model used in \cite{polanco2020} and the one used in our work describe superfluid helium at very different scales.
%, hence triggering non-fully reconnections. 
Our work hence motivates further detailed studies of
the role played by helicity in superfluid dynamics 
\cite{dileoni2016,galantucci2021},
moving the emphasis from few vortex systems \cite{scheeler2014} 
to fully coupled superfluid turbulence.
